---
title: "Machine Learning Final Project"
author: "Maddie Lee"
date: "2024-12-11"
output:
  html_document:
    number_sections: no
    toc: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: inline
execute:
  warning: no
  message: no
---

```{r message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary libraries
options(rgl.useNULL = TRUE)
library(dataPreparation)
library(doParallel)
registerDoParallel(cores = 4)
library(readxl)
library(mlbench)
library(glmnet) 
library(Rtsne)
library(Rtsne)
library(knitr)
library(caret)   
library(dplyr)   
library(rminer)
library(rmarkdown)
library(tidyverse) 
library(DescTools)
library(ggplot2)
library(lubridate)
library(tidymodels)
library(Ckmeans.1d.dp)
library(rpact)
library(neuralnet)
library(MatchIt)
library(marginaleffects)
library(quickmatch)
library(kableExtra)
library(car)
library(vip)
library(corrplot)
library(ROCR)
library(psych)
library(faraway)
library(neuralnet)
library(mltools)
library(lmtest)
library(tseries)
library(mclust)
library(forecast)
library(reshape2)
library(kernlab)
library(e1071)
library(xgboost)
library(yardstick)
library(tune)
library(randomForest)
library(ggcorrplot)
library(tidyverse)
library(FactoMineR)  
library(factoextra)
library(cluster)   
library(gridExtra)   
library(NbClust)     
```


# Introduction

__Perspective:__ \ 

Since 2020, I have been working in the mental health field, specifically in residential treatment for at-risk youth. I've always been passionate about mental health and hope to find a role that combines this passion with my skills and expertise in analytics and statistics. This assignment provided a great opportunity to explore mental health datasets and examine the risk factors that may influence overall mental well-being.

For this analysis, I assumed the perspective of a data scientist working for a healthcare organization that specializing in mental health treatment. The goal is to better understand factors influencing mental health outcomes among diverse individuals and provide data-driven insights for targeted interventions. Insights from this analysis can guide mental health advocates and healthcare providers in developing programs that address common stressors and promote mental well-being.

__Question:__ \ 

Can we predict whether an individual reports a mental health condition based on demographic and lifestyle factors using different predictive models?

__Models:__ \ 

- Penalized Regression
- SVM
- Ensemble Methods - Boosting
- Ensemble Methods - Bagging
- Neural Networks 

__Key Contributions:__\

1. **Support Vector Machines (SVM)**  
   SVM demonstrated limited predictive power with a testing accuracy of 0.45. Its contribution was minimal as it struggled to generalize patterns in the data.

2. **Penalized Regression (LASSO)**  
   LASSO provided stable but moderate performance with consistent accuracy of 0.515. It effectively identified key predictors such as age, work hours, and country through feature selection.

3. **Boosting**  
   Boosting mirrored LASSO’s results with consistent testing accuracy of 0.515, highlighting the robustness of the identified predictors without offering notable improvements.

4. **Bagging**  
   Bagging achieved perfect training accuracy (1.0) but testing accuracy of 0.495 revealed overfitting, limiting its usefulness for generalization.

5. **Neural Networks**  
   Neural networks performed best on the training data (accuracy of 0.735) but showed moderate testing accuracy (0.515), indicating overfitting. Despite this, neural networks captured complex, non-linear relationships, complementing insights from LASSO.


__Conclusions:__\

Did all methods support your conclusions or did some provide conflicting results? If so they provided conflicting results, how did you reconcile the differences?

Most methods, particularly LASSO, boosting, and neural networks, supported the primary conclusions by consistently identifying age, work hours, and country as key predictors. However, SVM and bagging showed less reliable results due to overfitting and lower testing accuracy.  

To reconcile these differences, the analysis focused on models with consistent testing performance and robust feature selection, such as LASSO and neural networks. This ensured the findings were validated by methods demonstrating generalizability, providing confidence in the conclusions.


__Data Source and Description:__ \ 

Source: [Kaggle](https://www.kaggle.com/datasets/bhadramohit/mental-health-dataset/data)

The dataset consists of simulated mental health information for 1000 individuals across various countries, professions, and lifestyles. 

```{r data table, echo=FALSE}

# Create a data frame with variable descriptions
data_description <- data.frame(
  Variable = c("Mental_Health_Condition", "Age", "Gender", "Occupation", "Country", 
               "Severity", "Consultation_History", "Stress_Level", 
               "Sleep_Hours", "Work_Hours", "Physical_Activity_Hours"),
  Type = c("Binary", "Numeric", "Categorical", "Categorical", "Categorical", 
           "Categorical", "Categorical", "Categorical", 
           "Numeric", "Numeric", "Numeric"),
  Description = c("Target variable: presence of mental health condition (Yes, No)",
                  "Age of the individual",
                  "Gender of the individual (Male, Female, Non-binary)",
                  "Occupation of the individual (IT, Healthcare, Education, etc.)",
                  "Country of residence (USA, India, UK, Canada, Australia)",
                  "Severity of mental health condition (None, Low, Medium, High)",
                  "Consultation history with a mental health professional (Yes, No)",
                  "Stress level (Low, Medium, High)",
                  "Average hours of sleep per day",
                  "Average hours of work per week",
                  "Average physical activity hours per week")
)

# Render the table with kable and add styling with kableExtra
kable(data_description, 
      caption = "Table 1: Data Description", 
      col.names = c("Variable", "Type", "Description"),
      align = "l") %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "grey") %>%  # Header styling
  column_spec(1, bold = TRUE) %>%  # Bold the first column
  column_spec(2, width = "10em") %>%  # Set column width for "Type"
  column_spec(3, width = "25em")  # Set column width for "Description"

```

# Read Data
```{r load datasets, warning=FALSE}
# Load data sets and clean variables
cloud_wd <- "/Users/madelinelee/Downloads"
setwd(cloud_wd)

mh <- read.csv("mental_health_dataset.csv")

# Convert categorical variables to factors
mh <- mh %>%
  mutate(across(c(Gender, Occupation, Country, Mental_Health_Condition, 
                  Severity, Consultation_History, Stress_Level), as.factor))

# Remove ID column for analysis
mh <- mh %>%
  select(-User_ID)

# Check for missing values
colSums(is.na(mh))


head(mh)
str(mh)
```

# EDA

## Summary Statistics for Numeric and Categorical Variables
```{r summary stats}
# Summary for numeric variables
summary(mh %>% select(Age, Sleep_Hours, Work_Hours, Physical_Activity_Hours))

# Summary for categorical variables
mh %>%
  select(Gender, Occupation, Country, Mental_Health_Condition, Severity, Consultation_History, Stress_Level) %>%
  summarise_all(~n_distinct(.))

```

## Distribution of Numerical Variables

```{r numeric dist}
# Histograms for numeric variables
ggplot(mh, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "white") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency")

ggplot(mh, aes(x = Sleep_Hours)) +
  geom_histogram(binwidth = 1, fill = "green", color = "white") +
  labs(title = "Distribution of Sleep Hours", x = "Sleep Hours", y = "Frequency")

ggplot(mh, aes(x = Work_Hours)) +
  geom_histogram(binwidth = 5, fill = "orange", color = "white") +
  labs(title = "Distribution of Work Hours", x = "Work Hours", y = "Frequency")

ggplot(mh, aes(x = Physical_Activity_Hours)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "white") +
  labs(title = "Distribution of Physical Activity Hours", x = "Physical Activity Hours", y = "Frequency")

# Boxplots for numeric variables
ggplot(mh, aes(y = Age)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Boxplot of Age", y = "Age")

ggplot(mh, aes(y = Sleep_Hours)) +
  geom_boxplot(fill = "green") +
  labs(title = "Boxplot of Sleep Hours", y = "Sleep Hours")

ggplot(mh, aes(y = Work_Hours)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Work Hours", y = "Work Hours")

ggplot(mh, aes(y = Physical_Activity_Hours)) +
  geom_boxplot(fill = "purple") +
  labs(title = "Boxplot of Physical Activity Hours", y = "Physical Activity Hours")

```


## Distribution of Categorical Variables

```{r}
# Bar plots for categorical variables
ggplot(mh, aes(x = Gender)) +
  geom_bar(fill = "lightblue") +
  labs(title = "Distribution of Gender", x = "Gender", y = "Count")

ggplot(mh, aes(x = Occupation)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Distribution of Occupation", x = "Occupation", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(mh, aes(x = Stress_Level)) +
  geom_bar(fill = "gold") +
  labs(title = "Distribution of Stress Level", x = "Stress Level", y = "Count")

```

## Target Distribution and Balance
```{r target distrubution}
ggplot(mh, aes(x = Mental_Health_Condition)) +
  geom_bar(fill = "grey") +
  labs(
    title = "Distribution of Mental Health Condition",
    x = "Mental Health Condition",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

mh %>%
  group_by(Mental_Health_Condition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)


```

Data appears balanced with 48.5% reporting no mental health condition and 51.5% reporting a mental health condition. This means we do not need to adjust for imbalanced data with up or downsampling with training the model.

## Correlation Analysis

```{r numeric correlation}
numeric_vars <- mh %>% select(Age, Sleep_Hours, Work_Hours, Physical_Activity_Hours)
correlation_matrix <- cor(numeric_vars)
ggcorrplot(correlation_matrix, lab = TRUE)
```

None of the numeric predictors have a strong correlation (all below +- .05)

## Bivariate Analysis
```{r bivariate}
# Boxplots to see how numeric variables relate to mental health condition
ggplot(mh, aes(x = Mental_Health_Condition, y = Age, fill = Mental_Health_Condition)) +
  geom_boxplot() +
  labs(title = "Mental Health Condition vs Age", x = "Mental Health Condition", y = "Age")

ggplot(mh, aes(x = Mental_Health_Condition, y = Sleep_Hours, fill = Mental_Health_Condition)) +
  geom_boxplot() +
  labs(title = "Mental Health Condition vs Sleep Hours", x = "Mental Health Condition", y = "Sleep Hours")

ggplot(mh, aes(x = Mental_Health_Condition, y = Work_Hours, fill = Mental_Health_Condition)) +
  geom_boxplot() +
  labs(title = "Mental Health Condition vs Work Hours", x = "Mental Health Condition", y = "Work Hours")

ggplot(mh, aes(x = Mental_Health_Condition, y = Physical_Activity_Hours)) +
  geom_boxplot(fill = "grey") +
  labs(
    title = "Mental Health Condition vs Physical Activity Hours",
    x = "Mental Health Condition",
    y = "Physical Activity Hours"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

There seems to be no major differences in numeric predictors between individuals that did and did not report a mental health condition.

# Data Splitting

```{r data split - 1}
# Set a seed for reproducibility
set.seed(123)

# Use createDataPartition to create a training index
train_index <- createDataPartition(mh$Mental_Health_Condition, p = 0.8, list = FALSE)

# Subset data into training and testing sets
train <- mh[train_index, ]
test <- mh[-train_index, ]

# Display the sizes of each dataset
cat("Training set size:", nrow(train), "\n")
cat("Testing set size:", nrow(test), "\n")

train_outcome <- train[, 5]
test_outcome <- test[, 5]

# Create dummy variables for categorical predictors
train_dummy <- dummyVars("~ .", data = train[, -5])
train_predictor <- predict(train_dummy, newdata = train[, -5])
test_predictor <- predict(train_dummy, newdata = test[, -5])

# Scale the predictors
train_predictor_scaled <- scale(train_predictor)
test_predictor_scaled <- scale(test_predictor, 
  center = attr(train_predictor_scaled, "scaled:center"), 
  scale = attr(train_predictor_scaled, "scaled:scale"))

# Convert outcome to binary (1 for "yes", 0 for "no")
train_outcome <- ifelse(train_outcome == "Yes", 1, 0)
test_outcome <- ifelse(test_outcome == "Yes", 1, 0)
```

# SVM

Scale

```{r scale}
# Scale the predictors after creating the recipe
train_predictor_scaled <- scale(train_predictor)
test_predictor_scaled <- scale(test_predictor, 
  center = attr(train_predictor_scaled, "scaled:center"), 
  scale = attr(train_predictor_scaled, "scaled:scale"))
```


Train

```{r train}
svm_model <- svm(
  x = train_predictor_scaled,
  y = as.factor(train_outcome),
  kernel = "linear",  # Using a linear kernel for simplicity
  cross = 5            # 5-fold cross-validation
)
# Print the SVM model summary
print(summary(svm_model))
```



Predict on Train

```{r predict train}
# Predict on the training set with SVM
y_train_pred_svm <- predict(svm_model, newdata = train_predictor_scaled)
train_accuracy_svm <- mean(y_train_pred_svm == train_outcome)
```

Predict on Test

```{r predict test}
# Predict on the test set with SVM
y_pred_svm <- predict(svm_model, newdata = test_predictor_scaled)
accuracy_svm <- mean(y_pred_svm == test_outcome)


cat("SVM Model\n")
cat("Training Accuracy:", round(train_accuracy_svm, 3), "\n")
cat("Testing Accuracy:", round(accuracy_svm, 3), "\n\n")
```
# Penalized Regression

Train Model
```{r penalized regression model}
# Train a penalized regression model (LASSO)
lasso_model <- cv.glmnet(
  x = as.matrix(train_predictor_scaled),
  y = train_outcome,
  alpha = 1,  # LASSO regression
  family = "binomial"  # Binary outcome
) 
```

Lambda
```{r best lamda}
# Print the best lambda value for LASSO
print(paste("Best lambda for LASSO:", lasso_model$lambda.min))

```

Predict on Train
```{r predict on train - pr}
# Predict on the training set with LASSO
y_train_pred_lasso <- predict(lasso_model, newx = as.matrix(train_predictor_scaled), s = "lambda.min", type = "class")
train_accuracy_lasso <- mean(y_train_pred_lasso == train_outcome)
```

Predict on Test
```{r predict on test - pr}
# Predict on the test set with LASSO
y_pred_lasso <- predict(lasso_model, newx = as.matrix(test_predictor_scaled), s = "lambda.min", type = "class")
accuracy_lasso <- mean(y_pred_lasso == test_outcome)

# Print results
cat("LASSO Model\n")
cat("Training Accuracy:", round(train_accuracy_lasso, 3), "\n")
cat("Testing Accuracy:", round(accuracy_lasso, 3), "\n\n")
```

# Ensemble Methods - Bagging

Model set up w/ Hyperparamter Tuning

```{r bagging}
# Define the Random Forest model specification with tunable hyperparameters
rf_spec <- rand_forest(
  trees = tune(),        # Number of trees
  mtry = tune(),         # Number of predictors to randomly sample
  min_n = tune()         # Minimum number of data points in a node
) %>%
  set_engine("randomForest") %>%
  set_mode("classification")

# Define the grid for hyperparameter tuning
rf_grid <- grid_regular(
  trees(range = c(200, 500)),
  mtry(range = c(2, round(sqrt(ncol(train))))),
  min_n(range = c(1, 20)),
  levels = 10
)

```


Cross-Validation and Hyperparameter Tuning


```{r recipe - bagging}
# Create a recipe for data preprocessing
rf_recipe <- recipe(Mental_Health_Condition ~ ., data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
```


```{r cv and tuning bagging}

# Set up cross-validation
folds <- vfold_cv(train, v = 5)

# Tune the model with grid search and cross-validation
rf_res <- tune_grid(
  rf_spec,
  preprocessor = rf_recipe,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc)
)

# Get the best tuned parameters
best_rf <- select_best(rf_res, metric = "roc_auc")


```

Train the Final Model w/ Best Parameters 

```{r train - bagging}
# Finalize the model with the best hyperparameters
final_rf <- finalize_model(rf_spec, best_rf)

# Train the model on the full training data
final_rf_fit <- final_rf %>%
  fit(Mental_Health_Condition ~ ., data = train)

# Make predictions on the training set and calculate accuracy
train_predictions <- predict(final_rf_fit, train) %>%
  bind_cols(train %>% select(Mental_Health_Condition))

train_accuracy_bag <- train_predictions %>%
  metrics(truth = Mental_Health_Condition, estimate = .pred_class) %>%
  filter(.metric == "accuracy") %>%
  .$`.estimate`


```

Evaluate the Model on the Test Set

```{r test - bagging}
# Make predictions on the test set
rf_predictions <- predict(final_rf_fit, test, type = "prob") %>%
  bind_cols(predict(final_rf_fit, test)) %>%
  bind_cols(test %>% select(Mental_Health_Condition))

# calculate accuracy
test_accuracy_bag <- rf_predictions %>%
  metrics(truth = Mental_Health_Condition, estimate = .pred_class) %>%
  filter(.metric == "accuracy") %>%
  .$`.estimate`

# Print results
cat("Bagging Model\n")
cat("Training Accuracy:", round(train_accuracy_bag, 3), "\n")
cat("Testing Accuracy:", round(test_accuracy_bag , 3), "\n\n")

```


# Ensemble Methods - Boosting

Regularization

Try adding regularization to prevent overfitting and improve model performance

```{r regularization}
xgb_spec_reg <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("xgboost", lambda = 1, alpha = 1) %>%
  set_mode("classification")

# Define the tuning grid
xgb_grid_reg <- grid_latin_hypercube(
  trees(range = c(50, 300)),
  tree_depth(range = c(3, 10)),
  learn_rate(range = c(0.01, 0.3)),
  loss_reduction(),
  sample_prop(range = c(0.6, 1.0)),
  finalize(mtry(), train),
  min_n(range = c(1, 10)),
  size = 20
)

# Create the recipe for data preprocessing
xgb_recipe <- recipe(Mental_Health_Condition ~ ., data = train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

# Set up 5-fold cross-validation on the training set
folds <- vfold_cv(train, v = 5)

# Tune the model with cross-validation
xgb_res_reg <- tune_grid(
  xgb_spec_reg,
  preprocessor = xgb_recipe,
  resamples = folds,
  grid = xgb_grid_reg,
  metrics = metric_set(roc_auc)
)

# Select the best hyperparameters based on AUC
best_xgb_reg <- select_best(xgb_res_reg, metric = "roc_auc")

# Finalize the model specification with the best hyperparameters
final_xgb_spec_reg <- finalize_model(xgb_spec_reg, best_xgb_reg)

# Fit the final model on the entire training set
final_xgb_fit_reg <- final_xgb_spec_reg %>%
  fit(Mental_Health_Condition ~ ., data = train)

# Predict on the training set
train_predictions <- predict(final_xgb_fit_reg, train, type = "class") %>%
  bind_cols(train %>% select(Mental_Health_Condition))

train_accuracy_boost <- mean(train_predictions$.pred_class == train_predictions$Mental_Health_Condition)

# Predict on the test set
test_predictions <- predict(final_xgb_fit_reg, test, type = "class") %>%
  bind_cols(test %>% select(Mental_Health_Condition))

test_accuracy_boost <- mean(test_predictions$.pred_class == test_predictions$Mental_Health_Condition)

# Print results
cat("Boosting Model\n")
cat("Training Accuracy:", round(train_accuracy_boost, 3), "\n")
cat("Testing Accuracy:", round(test_accuracy_boost, 3), "\n\n")

```


# Neural Networks

Data Splitting

```{r data split}
# Set a seed for reproducibility
set.seed(123)

# Use createDataPartition to create a training index
train_index <- createDataPartition(mh$Mental_Health_Condition, p = 0.8, list = FALSE)

# Subset data into training and testing sets
train <- mh[train_index, ]
test <- mh[-train_index, ]

# Display the sizes of each dataset
cat("Training set size:", nrow(train), "\n")
cat("Testing set size:", nrow(test), "\n")

train_predictor <- train[,-5]
train_outcome <- train [,5]
test_predictor <- test[,-5]
test_outcome <- test [,5]
```

Cross Validation
```{r cv}
ctrl <- trainControl(method = "repeatedcv", # cross-validation
  number = 5, # 5 folds
  repeats = 5,
  classProbs = TRUE # report class probability
)
```

NN Model
```{r tuning}
avnnetGrid <- expand.grid(decay = c(.001,.005),
  size = c(15:35),
  bag = FALSE)

avnnet_fit <- train(train_predictor, train_outcome,
  method = "avNNet",
  tuneGrid = avnnetGrid,
  trControl = ctrl,
  softmax = TRUE,
  reProc = c("center", "scale"),
  trace = FALSE,
  metric= "Accuracy",
  allowParallel=TRUE,
  maxit = 50)

plot(avnnet_fit)
```

Predicting on train and test
```{r predict}
# Predict on training data
train_predictions <- predict(avnnet_fit, train_predictor)
train_accuracy <- mean(train_predictions == train_outcome)

# Predict on testing data
test_predictions <- predict(avnnet_fit, test_predictor)
test_accuracy <- mean(test_predictions == test_outcome)

# Print results
cat("Nueral Network Model\n")
cat("Training Accuracy:", round(train_accuracy, 3), "\n")
cat("Testing Accuracy:", round(test_accuracy, 3), "\n")

```

Confusion Matrix

```{r confusion matrix}
# Generate confusion matrix for testing data
# Generate confusion matrix for training data
train_confusion_matrix <- table(Predicted = train_predictions, Actual = train_outcome)
test_confusion_matrix <- table(Predicted = test_predictions, Actual = test_outcome)
cat("Confusion Matrix - Training Data:\n")
print(train_confusion_matrix)

cat("\nConfusion Matrix - Testing Data:\n")
print(test_confusion_matrix)
```
Feature Importance

```{r feature importance}

# Function to calculate permutation importance
calculate_permutation_importance_avnnet <- function(model, test_X, test_y) {
  # Base predictions and accuracy
  base_predictions <- predict(model, test_X)
  base_accuracy <- mean(base_predictions == test_y)
  
  # Calculate importance for each feature
  feature_importances <- sapply(1:ncol(test_X), function(i) {
    # Permute one feature
    permuted_X <- test_X
    permuted_X[, i] <- sample(permuted_X[, i])
    
    # Predictions with permuted feature
    permuted_predictions <- predict(model, permuted_X)
    permuted_accuracy <- mean(permuted_predictions == test_y)
    
    # Importance as drop in accuracy
    base_accuracy - permuted_accuracy
  })
  
  names(feature_importances) <- colnames(test_X)
  return(feature_importances)
}

# Use your testing dataset for evaluation
feature_importances <- calculate_permutation_importance_avnnet(avnnet_fit, test_predictor, test_outcome)

# Sort and display feature importances
feature_importances <- sort(feature_importances, decreasing = TRUE)
print(feature_importances)

# Visualize feature importance with smaller text
barplot(
  feature_importances,
  names.arg = c("Age", "Country", "Work_Hours", "Physical_Activity_Hours", 
                "Gender", "Occupation", "Consultation_History", "Severity", 
                "Stress_Level", "Sleep_Hours"),
  main = "Feature Importance for avNNet Model",
  horiz = TRUE,
  col = "grey",
  las = 1,
  cex.axis = 0.6,
  cex.lab = .6,
  cex.names = 0.35,  # Reduce text size for y-axis labels
  xlab = "Importance (Drop in Accuracy)"
)

```


